Graphical Tests for Power Comparison of Competing Designs

### 2012-03-30 setup.tex
Merge branch 'master' of github.com:heike/comparing-designs
Conflicts:
	setup.tex
table 1 reorganized to match the plots

Two recent data analyses motivated the setup of this experiment. The first one arises from an analysis of US flight traffic data, and a finding related to how wind direction affects the efficiency of an airport.  The second one arose during the review of a paper claiming that it was impossible to statistically test for differences in center between two samples where sample size was small and from a non-normal population, with application to studying toxic waste sites. Visually, using lineups of dotplots, it was clear that there was a difference.

@@

@@

An additional factor we were interested in this particular situation was to assess, how much of the data we actually needed to use in a design to have observers pick out the pattern. Clearly, a design is more efficient, if a smaller sample size is sufficient for showing the presence of a relationship. In order to investigate this, we took different size samples of the data and created lineup plots of all four designs for these subsets. The effect of sampling size on the displays mostly stems from the additional variability introduced when using small samples, which might hide the pattern, if it is not displayed prominently. 
%XXX picture of sampling size?
Another perturbation to the data are `shifts' in wind direction, i.e. we make use of the circular nature of the wind direction and adjust the $x$ axis by using different offsets.

@@

\begin{table}[hbtp]
\resizebox{\linewidth}{!} {
	\rowcolors{1}{white}{lightgray}
	\begin{tabular}{ll|@{}r|@{}r|@{}r|@{}r|@{}r|@{}r}
	\multicolumn{2}{l}{sample size (in percent)}  & 2 & 4 & 6 & 8 & 10 & 24 \\ [1pt] \hline
	Polar & with & 13/49& 9/43 & 9/39 & 7/35 & 8/40 & 13/38 \\
	& without & 6/51&   4/34 &  5/34 &  6/39 & 11/43 &  12/37\\ [1pt] \hline
	Euclidean & with &12/24& 18/27 & 42/44 & 33/41 & 39/43 & 46/47\\
	& without & 13/41 &24/32& 39/44 & 47/51 & 40/45 & 34/37\\
	\end{tabular}
	}
\caption{\label{tbl:treatment} Breakdown of lineups: correct/shown. Each participant was shown eight lineups from the set above and two separate `test' charts.}
\end{table}

@@

\subsection{Comparing Distributions}

In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions. As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. For each of these combinations, we created three replicate data sets and four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allows for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}

>>>>

Two recent data analyses motivated the setup of this experiment. The first one arises from an analysis of US flight traffic data, and a finding related to how wind direction affects the efficiency of an airport.  The second one arose during the review of a paper claiming that it was impossible to statistically test for differences in center between two samples where sample size was small and from a non-normal population, with application to studying toxic waste sites. Using lineups containing dotplots we found it was possible to detect a difference, and we were curious to see if other display types (density, histogram, boxplot) might be better than dotplots.

@@

*** Null hypothesis for the lineup?
$Ho: $ wind direction has no effect on efficiency and $H_a:$ wind direction does affect efficiency.

@@

An additional factor we were interested in this particular situation was to assess, how much of the data we actually needed to use in a design to have observers pick out the pattern. Clearly, a design is more efficient, if a smaller sample size is sufficient for showing the presence of a relationship. In order to investigate this, we took different size samples of the data and created lineup plots of all four designs for these subsets. The effect of sampling size on the displays mostly stems from the additional variability introduced when using small samples, which might hide the pattern, if it is not displayed prominently.

Another perturbation to the data are `shifts' in wind direction, i.e. we make use of the circular nature of the wind direction and adjust the $x$ axis by using different offsets.

@@

\begin{table}[hbtp]
\resizebox{\linewidth}{!} {
	%\rowcolors{1}{white}{lightgray}
	\begin{tabular}{ll@{   }r@{   }r@{   }r@{   }r@{   }r@{   }r}
	& & \multicolumn{6}{c}{sample size}  \\
        \multicolumn{2}{c}{type of chart} & 2 & 4 & 6 & 8 & 10 & 24 \\ [1pt] \hline 
	euclid & with & 12/{\bf 24}& 18/{\bf 27} & 42/{\bf 44} & 33/{\bf 41} & 39/{\bf 43} & 46/{\bf 47}\\
	       &      &(0.50)& (0.67) & (0.95) & (0.80) & (0.91) & (0.98)\\[1pt]
	& without & 13/{\bf 41} &24/{\bf 32}& 39/{\bf 44} & 47/{\bf 51} & 40/{\bf 45} & 34/{\bf 37}\\
	&         & (0.32) &(0.75)& (0.89) & (0.92) & (0.89) & (0.92)\\[3pt] %\hline
	polar & with & 13/{\bf 49}& 9/{\bf 43} & 9/{\bf 39} & 7/{\bf 35} & 8/{\bf 40} & 13/{\bf 38} \\
	      &      & (0.27)& (0.21) & (0.23) & (0.20) & (0.20) & (0.34) \\[1pt]
	& without & 6/{\bf 51}&   4/{\bf 34} &  5/{\bf 34} &  6/{\bf 39} & 11/{\bf 43} &  12/{\bf 37} \\
        &         & (0.12) & (0.12) &  (0.15) &  (0.15) & (0.26) &  (0.32)\\ 
	\end{tabular}
	}
\caption{\label{tbl:treatment} Breakdown of lineups: number correct/\textbf{number shown} (proportion correct). Each participant was shown eight lineups, and the same lineup was shown to multiple people (denominator in the table) reasonably well-spread among the treatment levels. }
\end{table}

@@

\subsection{Comparing Distribution Centers}

In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions. The null hypothesis of each lineup is $H_o:$ centers of the two groups are the same, and the alternative hypothesis is $H_a:$ center of the blue group is shifted to the right.

As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. For each of these combinations, we created three replicate data sets and four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allows for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}
